I would like to thank my advisor Lev for his advice, support, and collaboration
throughout my studies. Additionally, I would like to thank the MCS group
faculty, especially Gy\"orgy Tur\'an, Dhruv Mubayi, Shmuel Friedland, as well
as the graduate students in our group, Sam Cole, Ben Fish, Yi Huang, \'Ad\'am
Lelkes, among others. I would further like to thank my summer hosts Chandrika
Kamath and Rajmonda Caceres for providing me with intellectually stimulating
summers. Also thanks to the many UIC faculty who inspired and
supported me during my indecisive years, including Brooke Shipley, Izzet
Coskun, David Dumas, and others. And to my fellow graduate students, who have
become close friends and whom there are too many to fit here. And finally,
thanks to my wife Erin and my family who encouraged and supported me in my
pursuits.

\vspace{1cm}
\noindent The work presented in this thesis was published in conferences and journals as the following:
\vspace{0.5cm}

\noindent Chapter 1~\cite{KunPR13}: Jeremy Kun, Brian Powers, Lev
Reyzin: Anti-coordination Games and Stable Graph Colorings. \textit{Symposium
on Algorithmic Game Theory (SAGT)} 2013: 122-133.

\noindent Chapter 2~\cite{KunR14}: Jeremy Kun, Lev Reyzin: On Coloring
Resilient Graphs. \textit{Mathematical Foundations of Computer Science (MFCS)}
(2) 2014: 517-528.

\noindent Chapter 3~\cite{FishKLRT15}: Benjamin Fish, Jeremy Kun,
\'Ad\'am D\'aniel Lelkes, Lev Reyzin, Gy\"orgy Tur\'an: On the
Computational Complexity of MapReduce. \textit{International Symposium on
Distributed Computing (DISC)} 2015: 1-15.

\noindent Chapter 4~\cite{GutfraindKLR15}: Alexander Gutfraind, Jeremy Kun,
\'Ad\'am D\'aniel Lelkes, Lev Reyzin: Network Installation Under Convex Costs.
\textit{Journal of Complex Networks} 2015 (in press).

Additional papers published by the author over the course of his
graduate studies are included in the Curriculum Vitae at the end of
this document.

